Finally Tagless, Partially Evaluated

### 2007-06-08 tagless/tagless-final.tex
added lost of comments, from the reviewers, etc. Go home to clean them up.

%\newcommand{\jacques}[1]{{\it [Jacques says: #1]}}
%\newcommand{\oleg}[1]{{\it [Oleg says: #1]}}
%\newcommand{\ccshan}[1]{{\it [Ken says: #1]}}

@@

@@

@@

We focus on embedding a typed object language into a typed metalanguage.
The benefit of types in this setting is to rule out meaningless object terms,
thus enabling faster interpretation and assuring that our interpreters
do not get stuck.
To be concrete, we use the typed object language in
Figure~\ref{fig:object} throughout this paper.  We aim not just for
evaluation of object programs but also for
compilation, partial evaluation, and other processing.

@@

%\oleg{see tagless\_interp1.ml, module Tagfull for the complete code.
%  If you change the code in here, please adjust the .ml file
%  accordingly. Let the paper and the accompanying code be in sync.}

@@

@@

@@

We have also implemented this compiler in Haskell. Since Haskell
has no (convenient, typed) staging facility, we had to emulate
it. To be precise, we defined a data type |ByteCode| with
constructors such as |Var|, |Lam|, |App|, |Fix|, and |INT|.
Whereas our representation of object terms uses HOAS,
our bytecode uses integer-named
variables to be realistic. We then define 
\begin{code}
newtype C t = C (Int -> (ByteCode t, Int)) 
\end{code}
where |Int| is the counter for creating fresh variable
names. We define the compiler by making |C| an instance of the
class |Symantics|. The implementation\footnote{The implementation uses
GADTs because we also wanted to write a typed interpreter for 
the \texttt{ByteCode} \emph{data type}.} is quite similar (but slightly more
verbose) than the corresponding MetaOCaml code above. The
accompanying code gives the full details.

@@

@@

@@

\acks%\subsection*{Acknowledgments}
We thank Martin Sulzmann and Walid Taha 
for helpful discussions.  Sam Staton, Pieter Hofstra, and Bart Jacobs
kindly provided some useful references on coalgebraic structures in the
$\lambda$\hyp calculus.

>>>>

\newcommand{\jacques}[1]{{\it [Jacques says: #1]}}
\newcommand{\oleg}[1]{{\it [Oleg says: #1]}}
\newcommand{\ccshan}[1]{{\it [Ken says: #1]}}

@@

% http://flint.cs.yale.edu/aplas2007/

@@

\oleg{Emphasize somewhere in the Intro that we solved the problem
exactly as posed in Walid's, Xi's, and SPJ's papers. And we obtain the
result in the form those papers declare desirable.
We use their running example and achieve the result they
call desirable.}

@@

We focus on embedding a \emph{typed} object language into a 
\emph{typed} metalanguage.
The benefit of types in this setting is to rule out meaningless object terms,
thus enabling faster interpretation and assuring that our interpreters
do not get stuck.
To be concrete, we use the typed object language in
Figure~\ref{fig:object} throughout this paper.  We aim not just for
evaluation of object programs but also for
compilation, partial evaluation, and other processing.

@@

% see tagless\_interp1.ml, module Tagfull for the complete code.
% If you change the code in here, please adjust the .ml file
% accordingly. Let the paper and the accompanying code be in sync.

@@

\oleg{We must stress that we do not Church-encode a universal type
(or the solution in the previous section). The latter will not be
patently total because we have to provide error continuatiuons and we
have to argue why they will not be called. A compiler is mots likely
incapable of such an argument and so has to compile them in. Here, we do
not have any error continuations at all.}

@@

\oleg{We must stress that: of course all parts of our solution are
known. That was the whole point. The problem was claimed to be
unsolvable with the existing means, and so new techniques have been
proposed. We showed that regardless of the merits, power and convenience of
those new techniques, the existing, old solutions were sufficient for
the stated problem -- thus correcting the public record (of ICFP and
POPL conferences). We don't say GADT, etc. are useless -- but we wish
to show to what extent the existing old methods suffice.}
\oleg{Our contribution that we make do with a very restricted polymorphism
(HM plus the module system) which guarantees inference in ML and
Haskell.}
\oleg{So, our thrust is: there is public (ICFP, POPL) record of the problem
being open and claims that solving it requires various
extensions. We show that the problem as formulated in the literature
can be solved with the existing means. We claim the novelty of this,
essentially negative, result. We specifically disclaim any new
inventions.}

Again, the fact that our
solution is elementary and uses only existing known tools to give a
previously unknown solution to an important problem should not be
counted as a disadvantage.
We did not claim to invent a new method (of partial evaluation or of
interpretation). Rather, we claim that existing, old methods, when
properly combined, can solve problem that was publicly declared to be
unsolvable with the existing, old means.
Some parts of our solution have been known before; we mention that in
the paper and reviewers point more references. However, none of the
parts have been combined to solve the problem stated as open in
the recent published literature Walid03 and WalidTagless.

@@

We have also implemented this compiler in Haskell. 
\oleg{mention TH solution in footnote. ``I was rather hoping it could
have been emulated via TemplateHaskell. I realise TH is untyped and a
compile-time staging system. The claim would have been that the TH
code produces will always be type correct. We could still map the
typed ByteCode data type to untyped TH abstract syntax.''}
Since Haskell
has no (convenient, typed) staging facility, we had to emulate
it. To be precise, we defined a data type |ByteCode| with
constructors such as |Var|, |Lam|, |App|, |Fix|, and |INT|.
Whereas our representation of object terms uses HOAS,
our bytecode uses integer-named
variables to be realistic. We then define 
\begin{code}
newtype C t = C (Int -> (ByteCode t, Int)) 
\end{code}
where |Int| is the counter for creating fresh variable
names. We define the compiler by making |C| an instance of the
class |Symantics|. The implementation\footnote{The implementation uses
GADTs because we also wanted to write a typed interpreter for 
the \texttt{ByteCode} \emph{data type}.} is quite similar (but slightly more
verbose) than the corresponding MetaOCaml code above. The
accompanying code gives the full details.

@@

\oleg{Rev3:In 4.3, please carefully distinguish the limitation of GHC's
implementation of GADTs from the type theory. The pattern match in 4.3
is statically exhaustive, and can be shown so in other systems. If
this code were implemented using inductive families Coq, for example,
the user would be able to (and in fact required) to prove that the
other cases could not occur.}

@@

Stress typed.

The new features may be deservingly
exciting and suitable. We do not wish to stop progress. We merely wish to set
the record straight as to which problem can be solved with which minimal and
widely available tools.

Even the idea of having multiple interpretations of the
pervasives of a standard language (as in the Symantics structure) has
been exploited by many (see e.g. the Lava and Hawk work, Augustsson's
clever two-level re-interpretation of Haskell (to generate residual
functions in Excel (!)), Rhiger's thesis, and many DSLs.

Stress differences from phantom types.
Use of phantom types is popular: see Lava 
Lava: Hardware Design in Haskell (1998)) and Rhiger's thesis.
(Lava: has universal type for Bit and NumSig and correspondent
tag matching: it is not tagless).

Rhiger's But Fig 2.2, p33: universal type Raw.  He uses phantom type
upon the Exp datatype. But that is cheating: phantom type means
essentially we can easily do coerce. We use real types.  That's why he
had to do tedious proofs in Sec 2 of soundness and completeness of
embedding. Whereas our proofs are obvious.  His sec 3 is based on data
representation of terms. They have type tags.  We do nothing of that
kind: See Sec 3.1.2. See numerous "data Term" in Sec3, which is the U
type.  In Sec 3.3.4 (p76) Rhiger specifically says that his encoding
cannot do typed CPS transformation -- whereas our does. BTW, Rhiger
thesis contains the definitions of the interpreter and the compiler,
in the beginning. Use this in response to Rev1)
OR:
The simplicity of our encoding enables trivial proofs of
soundness and completeness compared to Rhiger's.  It also makes
self-interpretation simple, which is a strength of our work.

first approach that gives typed CPS in a HM language. We also need to
cite the paper by Stefan Monnier at PLPV2006 where they say, I think,
that typed CPS is impossible without GADT or like.

the basic idea of this paper (the encoding of the term
GADT as its elimination form, i.e. as a coalgebraic structure) is not
new to this work. It appeared many years ago in: Metacircularity in
the polymorphic lambda-calculus, Frank Pfenning and Peter Lee.
Theoretical Computer Science 89 (1991) 137-159. (Other encodings of
GADTs have appeared since Pfenning and Lee's work, for example, in
work on generic programming.)
BTW,
many GADT encodings I've seen (Sulzmann's et al) require existentials,
which are not available in ML. The only GADT encodings I've seen in ML
use unsafe operations.

It seems that Pfenning and Lee embed $F_2$ with type constructions in 
(pure) $F_3$.  We embed $F_1$ in (weak?) $F_2$, as I see it.  In a way, what 
we do is very similar to what they do (Figure 1, p.152), except that we 
do it in standard programming languges.  It is unclear if their work can 
be implemented (yet) in any language.  And we preserve type-inference, 
while their solution needs explcit types!
The following line of their conclusion is worth citing: "... this does 
not imply that the same language is also suitable for type 
metaprogramming. ... such as partial evaluation... do not seem to be 
expressible".
I suspect you're right, but I'm still reading the paper.  See also page
146: "for a term M in $F_1$ (a simply-typed term), the representation
$\bar{M}$ will be in $F_2$".  The move from $F_2$ to $F_3$ 
and beyond reminds me
strongly of our attempts at self-interpretation without the notion of a
syntactic hole.
    Pfenning and Lee conclude that partial evaluation and program
    transformations "do not seem to be expressible" even using their
    extension to F-omega, perhaps because their avoidance of general
    recursive types compels them to include the polymorphic lift that we
    avoid in our Section 4.1.

 Although GADTs
may indeed be "more flexible and easier to use", they are not
available in mainstream ML, and GHC as released does not implement
them correctly (as Simon Peyton-Jones admits publicly).  Whereas
Pasalic et al. have already solved the problem of tagless typed
(staged) interpretation, our solution is the first to use only
facilities available in mainstream ML and Haskell, namely
Hindley-Milner and either simple modules or constructor classes.  The
simplicity of these facilities is again a strength of our work: it
makes embedded DSLs more practical to build.

Type-indexed types have been around since at least Harper and
Morrisett's Compiling Polymorphism Using Intensional Type Analysis,
POPL 1995. They also were supported by Generic Haskell (Hinze,
Juering, Loeh, Type Indexed Datatypes, SCP 2004).
Generic Haskell is not mainstream.  Xi et al. point out why
intensional type analysis cannot handle HOAS with fix.

@@

\oleg{From Rev3}This representation limited to primitive
recursive folds over the datatype (as that is the elimination form
provided.) That means that some operations are easy implement
(interpretation, size calculation, etc.), some can be done with effort
(one-pass cps, and other mutually recursive functions), and perhaps
some cannot be implemented at all (those that are not primitive
recursive). For example, can you implement a transformation that
eliminates eta-expansions from a term?

\acks%\subsection*{Acknowledgments}
We thank Martin Sulzmann and Walid Taha 
for helpful discussions.  Sam Staton, Pieter Hofstra, and Bart Jacobs
kindly provided some useful references on coalgebraic structures in the
$\lambda$\hyp calculus. We thank anonymous reviewers for pointing out
related work.
